Precision line-by-line forward models for \'echelle spectra

### 2022-08-11 paper/paper1/ms.tex
Add a bunch of text, probably too much, on conceivable science

\subsection{GPU and Autodiff specific considerations}

@@

\subsection{Line-by-line fundamental parameter estimates}
The conversion of spectral line depths to spectral type\citep{1901AnHar..28..129C}---and by extension $T_\mathrm{eff}$ \citep{1925PhDT.........1P}---has occupied a large chunk of stellar spectroscopy in the last century, and the measurement of fundamental stellar properties remains one of the principal applications of astronomical spectroscopy.  Blas\'e as it currently stands does not output fundamental properties, and in fact it takes them \emph{for granted}: you must specify \emph{e.g.} the $T_\mathrm{eff}$, $\log{g}$, [Fe/H], and [$\alpha$/Fe] of your PHOENIX template as Step Zero before that template gets cloned and warped to match data.  So obtaining a $T_\mathrm{eff}$ estimate is impossible.

Simply put, we seek a function, $f$:
$$f_j(a_j, \sigma_j, \gamma_j) \to (T_\mathrm{eff}, \log{g}, [\mathrm{Fe}/\mathrm{H}])$$

@@

Fascinatingly, there are two related ways forward towards this function.  They both involved ensembles of spectra, such as a spectral atlas or library, and they both involve first obtaining the inverse function:

\begin{eqnarray}
    g_j(T_\mathrm{eff}, \log{g}, [\mathrm{Fe}/\mathrm{H}]) &\to& (a_j, \sigma_j, \gamma_j)
\end{eqnarray}

A purely model-based approach would start by cloning every single PHOENIX spectrum in the grid dimensions-of-interest, over the wavelength range-of-interest.  The resulting product would be a line-by-line catalog of cloned spectral properties for each 3D coordinate on this grid.  We could then assemble a heatmap of for how each $j^{th}$ line properties changes across this heatmap, $g_j$.  In this way, we are reverse-engineering the \emph{e.g.} temperature dependence of each spectral line, as encoded by the PHOENIX atmosphere models.  The final step would involve finding the neareast neighbor of each $j^{th}$ observed spectral line to each $j^{th}$ grid heatmap point; that is, finding a way to invert the function $g_j$ to get $f_j$.  The information-weighted mean, median, or mode of these nearest neighbors could then be reported as a revised ``best-fit $T_\mathrm{eff}$'', for instance.

@@

The construction and calibration of these ensembles of spectra represents a tremendous amount of work beyond the scope of this paper.  But its inevitable creation represents a major breakthrough for the field of stellar astrophysics.  It would be an extremely precise, accurate, fast, and interpretable way to measure the properties of stars based solely on their high bandwidth \'echelle spectrum.

@@

@@

\texttt{blas\'e} offer a new approach to Doppler imaging that may overcome these historical limitations by simultaneously fitting both the imperfections in the underlying spectrum and its line profile perturbations.  This approach is analogous to the linearized model in \citet{2021arXiv211006271L}, but with the benefit of also handling non-linear properties of the spectrum such as line widths, shapes, and locations, while also handling telluric contamination.  Hypothetically the \citet{2021arXiv211006271L} approach could be partially absorbed into \texttt{blase}, or vice versa, though such a merger may be complicated to implement.

\subsection{Extreme Precision Radial Velocity}

Already \texttt{blas\'e} is equipped to fit every single line with its own systemic $RV$, by tuning the line center position $\lambda_\mathrm{c}$ at each training epoch.  There exists both empirical evidence and some theoretical motivation that $RV$ jitter varies from line-to-line \citep{2018A&A...620A..47D}.  Importantly, the extent of this line-by-line RV jitter could be predicted in part by the depth of line formation.  A future extreme-precision-RV (EPRV) version of \texttt{blase} could leverage this information.  As an example, the depth-of-formation for all (or a subset) of lines could be obtained and associated with each spectral line.  A regularization scaling term could be introduced to allow the line positions to vary, but only in proportion to their depth of formation.

\subsection{Circumstellar disk veiling}
To-be-completed

\subsection{Accretion Veiling}
To-be-completed

>>>>

\subsection{GPU and Autodiff specific considerations}\label{gpuConsiderations}

@@

\subsection{Extreme Precision Radial Velocity} \label{eprv}

Already \texttt{blas\'e} is equipped to fit every single line with its own systemic $RV$, by tuning the line center position $\lambda_\mathrm{c}$ at each training epoch.  There exists both empirical evidence and some theoretical motivation that $RV$ jitter varies from line-to-line \citep{2018A&A...620A..47D}.  Importantly, the extent of this line-by-line RV jitter could be predicted in part by the depth of line formation \citep{2021A&A...653A..43C, 2022A&A...664A..34A}.  A future extreme-precision-RV (EPRV) version of \texttt{blase} could leverage this information.  As an example, the depth-of-formation for all (or a subset) of lines could be obtained and associated with each spectral line.  A regularization scaling term could be introduced to allow the line positions to vary, but only in proportion to their depth of formation.

\subsection{Line-by-line fundamental parameter estimates}
The conversion of spectral line strengths to spectral type \citep{1901AnHar..28..129C}---and by extension line strengths to $T_\mathrm{eff}$ \citep{1925PhDT.........1P}---has occupied a large chunk of stellar spectroscopy in the last century.  The measurement of fundamental stellar properties remains one of the principal applications of stellar spectroscopy.  Blas\'e as it currently stands does not output fundamental properties, and in fact it takes them \emph{for granted}: you must specify \emph{e.g.} the $T_\mathrm{eff}$, $\log{g}$, [Fe/H], and [$\alpha$/Fe] of your PHOENIX template as Step Zero before that template gets cloned and warped to match data.  So obtaining a $T_\mathrm{eff}$ estimate from blas\'e is impossible, at least as disscussed so far.

The clearest way forward would be to calibrate the blas\'e-derived line properties.  Simply put, we seek a set of functions, $f$, that relate the line properties back to fundamental properties:
$$f_j(a_j, \sigma_j, \gamma_j) \to (T_\mathrm{eff}, \log{g}, [\mathrm{Fe}/\mathrm{H}])$$

@@

Fascinatingly, there are two related ways forward towards this function.  They both involve ensembles of spectra---such as a spectral atlas, spectral sequence, or library---and they both involve first obtaining the inverse function:

\begin{eqnarray}
    g_j &\equiv& f^{-1}_j\\
    g_j(T_\mathrm{eff}, \log{g}, [\mathrm{Fe}/\mathrm{H}]) &\to& (a_j, \sigma_j, \gamma_j)
\end{eqnarray}

A purely model-based approach would start by blas\'e-cloning every single PHOENIX spectrum in the grid dimensions-of-interest, over the wavelength range-of-interest.  The resulting product would be a line-by-line catalog of cloned spectral properties for each 3D coordinate on this grid.  We could then assemble a heatmap of for how each $j^{th}$ line properties changes across this heatmap, $g_j$.  In this way, we are reverse-engineering the \emph{e.g.} temperature dependence of each spectral line, as encoded by the PHOENIX atmosphere models.  The final step would involve finding the neareast neighbor of each $j^{th}$ observed spectral line to each $j^{th}$ grid heatmap point; that is, finding a way to invert the function $g_j$ to get $f_j$.  The information-weighted mean, median, or mode of these nearest neighbors could then be reported as a revised ``best-fit $T_\mathrm{eff}$'', for instance.

@@

The construction and calibration of these ensembles of spectra represents a tremendous amount of work beyond the scope of this paper.  But its creation could an extremely precise, fast, interpretable and reasonably accurate way to measure the properties of stars based solely on their high bandwidth \'echelle spectra, albeit with the overall accuracy hinging on the accuracy of the stellar benchmark labels.  The mechanics of this approach could be adapted to fit within existing pipelines such as SAPP \citep{2022A&A...658A.147G}.

There exists a corrolary from the method described here to the approach of \citet{czekala15}.  There, a 3D heatmap was created for relating the eigenweights of a PCA basis back to the stellar fundamental properties, for spectrum emulation purposes.  Here, the 3D heatmap is created on a line-by-line basis, and is therefore interpretable.  The PCA eigenspectra were generally uninterpretable, at least not easily.  So in the limiting case of obtaining a densely calibrated semi-empirical heatmap for all spectral lines across all grid dimensions, we will have achieved a spectral emulator that would obviate the need to clone spectra in the first place.  We would simply start with this powerful line-by-line emulator as a forward model, and go directly to the extrinsic warping and data-model comparison.

@@

Stellar abundance work involves precisely measuring line strengths (often reported as EWs) of different chemical constituents evinced in a star or substar's photosphere.  Relating those EWs back to physics can be done in a few ways.  Most easily, trends and patterns in the EWs can be assembled, and metal-rich and metal-poor clusters can be identified.  There exists a precision/accuracy tradeoff in the measurement process, usually stemming from the placement of the continuum, or assumptions about line blending.  Blas\'e offers an immediate solution to these challenges, since it starts from our best-guess for how nearby lines may be shaping the continuum.  Blas\'e appears to be a gateway to extremely fast and intervention-free way to obtain abundances, potentially useful for large surveys like APOGEE \citep{2017AJ....154...94M}, Gaia-ESO \citep{2012Msngr.147...25G}, RAVE \citep{2006AJ....132.1645S,}, and more.

This spirit resembles \texttt{ROBOSPECT} \citep{2013PASP..125.1164W}, and other approaches aimed at automatically measuring equivalent widths for a large number of stars.

One missing piece is either to post-process the blas\'e-output EW catalog with chemical labels (C, N, O, Si, Ba, Eu, etc.), or incorporate that information directly into the structure of blas\'e.  The latter case is intriguing.  For example, PHOENIX may not yet be aware of the existence of some obscure Lanthanum (La) line, possibly because the line was not yet understood and incorporated at the time of PHOENIX computation.  Rather than wait for PHOENIX to receive the necessary update, blas\'e could instantiate one or more such lines in the locations guided by the atomic line list.  This line could get initialized with some small amplitude, and would get added onto the underlying PHOENIX source as yet another freely tunable line.  In this way, all the other nearby lines would be used for assigning continuum and blending, but this free-standing Lanthanum line would acquire a value properly taking into account the impact of its neighbors.

@@

\texttt{blas\'e} offers a new approach to Doppler imaging that may overcome these historical limitations by simultaneously fitting both the imperfections in the underlying spectrum and its line profile perturbations.  This approach is analogous to the linearized model in \citet{2021arXiv211006271L}, but with the benefit of also handling non-linear properties of the spectrum such as line widths, shapes, and locations, while also handling telluric contamination.  Hypothetically the \citet{2021arXiv211006271L} approach could be partially absorbed into \texttt{blase}, or vice versa, though such a merger may be complicated to implement.

\subsection{Starspots and magnetic fields}
One current assumption of starspot spectral decomposition is that the starspot spectrum itself resembles the stellar photosphere of a cooler star.  This assumption appears adequate for detecting starspot spectra and measuring their physical properties \citep{gullysantiago17}.  But to second order starspots should exhibit some spectral peculiarities that make them depart from a ``normal'' stellar photosphere.  We may be probing deeper into the photosphere, and so the lines may experiencing higher pressure, with slightly different line widths.  Or maybe the finite convective velocity shift can be directly seen as systematic shifts of the spectral lines (as stated in Section \ref{eprv}).

Rather than applying fixed a mixture model of PHOENIX templates, one could adopt a mixture model of pre-cloned blas\'e models.  Then, the imperfections in the starspot spectrum can be learned alongside the filling factor of the starspots.

Often the most spotted stars are also rapidly rotating, and so the large $v\sin{i}$ blurs the blurs the starspot spectrum into the ambient photosphere.  One experiment could measure the starspot spectral peculiarities more easily: obtain multi-epoch high resolution spectra of a pole-on (or nearly pole-on) young star, and monitor it for secular variations attributable to stellar activity cycles.  Those cycles should confirm the ebb-and-flow of starspot area confirming the interpretation of two components as spots, and not binarity or some other anomaly.

Relatedly, magnetic fields information could be incorporated into blas\'e as Land\'e $g-$factors, to allow magnetic sensitive lines to exhibit either Zeeman broadening or Zeeman splitting.  This labelling could be conducted in the same way as described in the Abundances subsection.

\subsection{Circumstellar disk and accretion veiling}
Circumstellar disk veiling suppresses the strengths of all spectral lines \emph{en masse}, as the stellar photosphere gets outshined by a hot disk and/or envelope.

The bulk properties of that disk/envelope could be incorporated with a simple physical model: a black body of temperature $T_\mathrm{disk}$ and solid angle $\Omega_{disk}$ \citep{2018ApJ...862...85G}.  The spectral shape of the veiling would require sufficiently high bandwidth spectra, such as Xshooter \citep{2011A&A...536A.105V} or possibly IGRINS.  The extent of veiling would be fit alongside all the other stellar spectral lines.  An MCMC approach may outperform blas\'e in accuracy, since the choice of picking an underlying spectral template is degenerate with the derived veiling.

Accretion veiling can be treated in an analogous way, but with some more complications.  Emission lines can be easily incorporated as described in Section \ref{gpuConsiderations}.  These emission lines could be initialized with a line list of Hydrogen lines, forbidden lines, and other conspicuous features.  The shapes of those lines can be affected by winds and other physical phenomena, yielding a variety of physically interpretable line shapes \citep{2022arXiv220802940E}.  Blas\'e could incorporate those lineshapes as templates shared among several lines originating from similar physical enviroments, but scaled and shifted based on the details of each individual line's radiative transfer properties.
